---
title: "spatial decon for DSP data"
author: "Joseph J Zhao"
date: "4/5/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load packages

```{r setup, include=FALSE}
package.name=c(
  # data manipulation
  "tidyverse", "readr", "readxl", "dplyr", "tidyr", "lubridate", "tibble", "plyr", "stringr", "stringi", "gtools", "BiocManager", "GeomxTools",
  # parallel processing
  "doParallel", "parallel", "foreach", "doSNOW",
  # bioinformatics
  "DESeq2","preprocessCore", "umap", "corto", "GSVA", "GSEABase", "DGEobj.utils", "msigdbr", "SpatialDecon",
  # survival analysis
  "survminer", "survival", "rstpm2", "survRM2",
  # regression stuff
  "MASS", "splines", "DGEobj.utils", "DirichletReg",
  # misc
  "cluster", "ResourceSelection", "digitize", "reshape2", "reshape",
  # plotting
  "blandr", "pheatmap", "ggrepel", "ggplot2", "corrplot", "ggpubr", "gplots", "grid", "circlize", "ggbeeswarm", "see"
)

for (package.name in package.name){
  tryCatch(
  {
  if (!require(package.name, character.only = TRUE)){ 
    install.packages(package.name, character.only = TRUE) 
    library(package.name, character.only = TRUE)} else {library(package.name, character.only = TRUE)}
  }, 
    error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

```

## Set main working directory

```{r working directory}
wd="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/"
setwd(wd)

dir.create(paste(wd,"output/", Sys.Date(), sep=""))
setwd(paste(wd, "output/",Sys.Date(), sep="") )

```

## load dataset and reference

```{r}

path="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/submissions/substudies/crcpm/submissions/ccr/r1/workings"

dsp_crc=readRDS(paste0(path, "/data/dsp/dsp_crc.rds"))

# load into lists
l_dsp=list("colorectal"=dsp_crc)
l_rna=list("colorectal"=dsp_crc@assayData$q_norm)
l_ref=list("colorectal"=pData(dsp_crc))

```

# SpatialDecon: Impute immune cell levels
original paper
https://www.nature.com/articles/s41467-022-28020-5
https://bioconductor.riken.jp/packages/devel/bioc/vignettes/SpatialDecon/inst/doc/SpatialDecon_vignette_NSCLC.html

## Load reference dataframe

```{r}

data("safeTME")
head(safeTME)
data("safeTME.matches")
head(safeTME.matches)

```

## running basic spatial decon

```{r}

l_spatialdecon=list()

for (i.gccrc in names(l_dsp)){
  
  res = runspatialdecon(object = l_dsp[[i.gccrc]],
                      norm_elt = "q_norm",
                      raw_elt = "exprs_batch",
                      X = safeTME,
                      align_genes = TRUE)
  
  l_spatialdecon[[i.gccrc]]=res
  
}

heatmap(t(res$beta), cexCol = 1, cexRow = 0.7, margins = c(10,7))

```

## running advanced spatial decon

```{r}

l_spatialdeconad=list()

for (i.gccrc in names(l_dsp)){
  
  restils = runspatialdecon(object = l_dsp[[i.gccrc]],
                            norm_elt = "q_norm",                    # normalized data
                            raw_elt = "exprs_batch",                      # expected background counts for every data point in norm
                            X = safeTME,                            # safeTME matrix, used by default
                            cellmerges = safeTME.matches,           # safeTME.matches object, used by default
                            cell_counts = l_dsp[[i.gccrc]]$nuclei,           # nuclei counts, used to estimate total cells
                            is_pure_tumor = grepl("tumor", l_ref[[i.gccrc]]$group),          # identities of the Tumor segments/observations
                            n_tumor_clusters = 5)                   # how many distinct tumor profiles to append to safeTME
  
  l_spatialdeconad[[i.gccrc]]=restils
  
}

heatmap(sweep(restils@experimentData@other$SpatialDeconMatrix, 1, apply(restils@experimentData@other$SpatialDeconMatrix, 1, max), "/"),
         labRow = NA, margins = c(10, 5))

data("cellcols")

# export

# saveRDS(restils, file=paste0(wd, "data/dsp_gc/data/dsp_spatialdecon_gc.RData"))

```

plot advanced results - prop bar plot

```{r fig.height=7, fig.width=10}

o = hclust(dist(t(restils$cell.counts$cell.counts)))$order
layout(mat = (matrix(c(1, 2), 1)), widths = c(7, 3))
TIL_barplot(t(restils$cell.counts$cell.counts[, o]), draw_legend = TRUE, 
            cex.names = 0.5)

for (i.gccrc in names(l_spatialdeconad)){
  restils=l_spatialdeconad[[i.gccrc]]
  restils$prop_of_all = replace(restils$prop_of_all, is.na(restils$prop_of_all), 0)
  o = hclust(dist(restils$prop_of_all))$order
  TIL_barplot(t(restils$prop_of_all[o,]), draw_legend = TRUE, cex.names = 0.01, main=i.gccrc)
}

```

### Heatmap of cell propotions

```{r fig.height=5, fig.width=15}

for (i.gccrc in names(l_spatialdeconad)){

  restils=l_spatialdeconad[[i.gccrc]]
  
  colMeta=l_ref[[i.gccrc]][, c("region", "class")]
  colMeta$region[colMeta$region=="peritoneal-tumor"]="peritoneal-metastasis"
  
  my_colour=list(
    "region"=c("peritoneal-metastasis"="firebrick3", "primary-normal"="goldenrod3", "primary-tumor"="dodgerblue3"),
    "class"=c("stroma"="green3", "tumor"="steelblue3")
  )
  
  # print heatmap
  
  mat=restils$prop_of_all
  mat[is.na(mat)]=0
  
  heat=pheatmap(t(mat),
                col=brewer.pal(n=10, "YlGnBu"),
                # scale="column",
                cluster_cols = T,
                clustering_distance_cols = "euclidean", 
                # clustering_method = "ward.D2",
                cluster_rows = T,
                annotation_col=(colMeta),
                show_colnames = F,
                show_rownames = T,
                annotation_names_col = T,
                cellwidth = 1,
                cellheight = 10,
                border_color=NA,
                annotation_colors = my_colour
              )
  
  pdf(paste(Sys.Date(),i.gccrc, "dsp_immunedeconprop_heatmap.pdf"), height=5, width=13)
  print(heat)
  dev.off()
}

```


## melt to long form

```{r}

l_immunedecon_melt=list()
l_immunedecon_avg=list()

for (i.dsp in names(l_spatialdeconad)){
  df_melt=reshape2::melt(as.matrix(l_spatialdeconad[[i.dsp]]$prop_of_all))
  colnames(df_melt)[1:2]=c("id", "var")
  pData(l_dsp[[i.dsp]])$id=rownames(pData(l_dsp[[i.dsp]]))
                                   
  df=left_join(df_melt, pData(l_dsp[[i.dsp]]), by="id")
  df$var=as.character(df$var)
  l_immunedecon_melt[[i.dsp]]=df
  
  # df$patientid=ifelse(grepl("SC", df$patientid), str_extract(df$patientid, "SC\\d{3}"), df$patientid)
  
  l_immunedecon_avg[[i.dsp]]=df %>% group_by(var, id, region, class) %>% summarise(value=mean(value), .groups = 'drop')
}

```

## export data

```{r}

saveRDS(l_immunedecon_melt, paste0(path, "/data/dsp/immunedecon/df_immunedecon_crc.rds"))

```


## Dirichlet regression for cell proportions across all comparisons (with ROI)
- focus on PM PT comparisons

```{r}

# define groups - as per manual order (put normal primary stroma at the back as controls)
groups=c("peritoneal-metastasis-stroma", "peritoneal-metastasis-tumor", "primary-tumor-tumor", "primary-tumor-stroma", "primary-normal-stroma") %>%  rev
combs=combinations(length(groups), 2)

l_dsp_dirich=list()

for (i.gccrc in names(l_ref)){
 
  l_ref[[i.gccrc]]$sample=rownames(l_ref[[i.gccrc]])
  restils=l_spatialdeconad[[i.gccrc]]
  
  pdf(paste(Sys.Date(), i.gccrc, "dsp_immunedecon_dirchlolipop.pdf"), height=3.2, width=6)
  
  # for (i.gccrc in unique(df_ref$gccrc)){
  for (i.combs in 1:nrow(combs)){
      g1=groups[combs[i.combs, 1]]
      g2=groups[combs[i.combs, 2]]
    
      df_temp=restils$prop_of_all %>% data.frame
      df_temp$sample=rownames(df_temp)
      
      df_temp=left_join(df_temp, l_ref[[i.gccrc]][,c("group", "sample")], by="sample")
      df_temp=subset(df_temp, 
                    (grepl(g1, df_temp$group) | grepl(g2, df_temp$group)))
      df_temp$group=factor(df_temp$group, levels=c(g1, g2))
      
      # prepare dirichlet regression model
      AL <- DR_data((df_temp[,1:14]))
      
      mod=DirichReg(AL ~ group, df_temp)
      sum.mod=summary(mod) 
      
      mat=sum.mod$coef.mat
      mat=subset(mat, !rownames(mat)=="(Intercept)") %>% data.frame
      
      # cleaning
      mat$cell=str_replace_all(colnames(df_temp)[1:14], "\\.", " ")
      colnames(mat)=c("coef", "se", "z", "pvalue", "cell")
      mat$padj=p.adjust(mat$pvalue, method="bonferroni")
      mat$padj[is.na(mat$padj)]=1
      rownames(mat)=NULL
      
      # cleaning data
      mat$sig=ifelse(mat$padj<0.05, "p<0.05", "not sig")
      mat$diffexp=ifelse(mat$coef<0, paste0("upregulated in ", g1), paste0("upregulated in ", g2))
      
      mat=mat[order(mat$coef), ]
      mat$cell=factor(mat$cell, levels = mat$cell)
      
      # throw into list
      l_dsp_dirich[[i.gccrc]][[paste(g2, " vs ", g1)]]=mat
      
      # lollipop plot
      plot_dirich_dotplot=ggplot()+
        geom_segment(data=mat, aes(x=cell, y=coef, xend=cell, yend=0, alpha=sig), size=1)+
        theme(legend.position = "top") +
        theme(axis.text.x=element_text(angle=90, hjust=1)) +
        geom_point(data=mat, aes(x=cell, y=coef, alpha=sig, col=diffexp, size=abs(coef)), stroke=0) +
        labs(y=paste0("dirichlet regression coefficient"), x="", col="", alpha="") +
        scale_colour_manual(values=c("firebrick3", "dodgerblue3")) +
        geom_hline(yintercept=0, color = "black") +
        coord_cartesian(ylim=c(-1,1)) +
        scale_alpha_manual(values=c(0.25,1)) +
        scale_size(range = c(4,7)) +
        guides(size = "none")
      
      plot_dirich_dotplot %>% print
    
}

dev.off()
  
}

```

### Manual scatter plot comparisons for correlation

```{r}

for (i.gccrc in names(l_dsp_dirich)[2]){
  
  # prepare dataset
  names(l_dsp_dirich[[i.gccrc]])
  
  df_temp1=l_dsp_dirich[[i.gccrc]]$`primary-tumor-tumor  vs  primary-tumor-stroma` # define comparison 1
  df_temp2=l_dsp_dirich[[i.gccrc]]$`peritoneal-metastasis-stroma  vs  peritoneal-metastasis-tumor` # define comparison 2
  df_temp2$coef=-df_temp2$coef
  
  rownames(df_temp1)=df_temp1$cell
  rownames(df_temp2)=df_temp2$cell
  
  colnames(df_temp1)=paste0(colnames(df_temp1), "_1")
  colnames(df_temp2)=paste0(colnames(df_temp2), "_2")
  
  # merge
  df_temp_comb=merge(df_temp1, df_temp2, by="row.names")
  
  # pearson correlation
  cor=cor.test(df_temp_comb$coef_1, df_temp_comb$coef_2)
  
  pdf(paste(Sys.Date(), i.gccrc, paste0("dsp_dirch_scatter_stroma.pdf")), height=4.5, width=5)
  
  p=ggplot(data=df_temp_comb, aes(x=coef_1, y=coef_2, alpha=(padj_1<0.05 | padj_2<0.05), col=diffexp_1,  label=cell_1)) +
    geom_smooth(method=lm, color="grey20", se=T, alpha=0.3, fill="#69b3a2", linewidth=0.5)+
    geom_point(stroke=NA, aes(size=abs(coef_1))) +
    # scale_size_manual(values=c(6:8)) +
    coord_cartesian(ylim=c(-2.5, 1), xlim=c(-2.5,1))+
    scale_alpha_manual(values=c(0.6,1))+
    scale_color_manual(values=c("goldenrod3", "firebrick3"))+
    geom_vline(xintercept=0) +
    geom_hline(yintercept=0) +
    theme(legend.position = "none")+
    geom_text_repel(min.segment.length = 0.005, box.padding = 0.5, size=3.5, max.overlaps = 20, color="black") +
    # coord_cartesian(ylim=c(-2,2), xlim=c(-2,8)) +
    labs(x="stroma                                 tumor\ndirichlet regression coefficient\nprimary tumor", y="peritoneal metastasis\ndirichlet regression coefficient\nstroma                                  tumor", title="Spatially resolved changes in immune composition")+
    annotate("text", hjust=0 , x=min(df_temp_comb$coef_1),  y=1.5, label=paste0("Pearson, r=", format(round(cor$estimate[[1]],3), nsmall=3), ", ",
         ifelse(cor$p.value<0.001, "p<0.001", paste0("p=", format(round(cor$p.value[[1]],3), nsmall=3)))) ) +
    geom_abline(linetype=2)
  
  print(p)
  
  dev.off()
  
}


```


## focus on PT PM tumor ROI comparisons

```{r}
path="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/submissions/substudies/crcpm/submissions/ccr/r1/workings"

dsp_crc=readRDS(paste0(path, "/data/dsp/dsp_crc.rds"))

crcpm_dspstromalclusters <- read_csv(paste0(path, "/data/dsp/crcpm_stromalclusters/crcpm_dspstromalclusters_results.csv"))[-1]
pData(dsp_crc)=left_join(pData(dsp_crc), crcpm_dspstromalclusters, by="dspid")

df_immunedecon=readRDS(paste0(path, "/data/dsp/immunedecon/df_immunedecon_crc.rds"))
df_immunedecon$dspid=str_replace_all(str_replace_all(df_immunedecon$id, ".dcc", ""), "-", "_")

df_immunedecon=left_join(df_immunedecon, crcpm_dspstromalclusters, by="dspid")
df_immunedecon$stromal_cluster[is.na(df_immunedecon$stromal_cluster)]="tumor"

df_immunedecon$var=str_replace_all(df_immunedecon$var, "\\.", " ")

```

```{r}

df_immunedecon$region=factor(df_immunedecon$region, levels=c("primary-tumor", "peritoneal-metastasis"))

ggplot(aes(x=region, y=value, color=region), data=subset(df_immunedecon, df_immunedecon$class=="tumor" & grepl("Treg|T cells", df_immunedecon$var))) +
  geom_point(position = position_jitterdodge(0.1, dodge.width = .1))+
  stat_n_text()+
  guides(color="none")+
  scale_color_manual(values=c("dodgerblue3", "firebrick3"))+
  labs(x="", y="proportion") +
  stat_summary(fun = "mean", size = 3, shape=95, color="black") +
  facet_wrap(~var, scale="free_y") +
  stat_compare_means(method="t.test")

```

