---
title: "crcpm A5 GSEA"
author: "Joseph J Zhao"
date: "9/18/2021"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---
# GSEA 
General steps
1. Calculate fold change values using the DESeq2 package (done, just import)
2. Throw the fold change values into GSEA
3. Export the GSEA enrichment scores into a table/dataframe
4. Export plots (figure a way to combine)

# Setup and working directory

```{r set up}
package.name=c(# GSEA analysis
  "clusterProfiler", "enrichplot", "pathview", "msigdbr", "fgsea", "gtools", "DOSE",
  # bioinformatics
  "DESeq2", "GSVA",
  # data manipulation
  "tidyverse", "reshape", "readr", "readxl", "dplyr", "tidyr", "lubridate", "tibble", "plyr", "devtools", "stringr", "stringi", "rlist",
  # parallel processing
  "doParallel", "parallel", "foreach", "doSNOW",
  # survival analysis
  "survminer", "survival", "rstpm2", "survRM2",
  # regression stuff
  "MASS", "splines", "Hmisc", "rms","twang",
  # misc
  "cluster", "ResourceSelection", "digitize", "tableone", "pROC",
  # plotting
  "blandr", "pheatmap", "ggrepel", "ggplot2", "corrplot", "ggpubr", "gplots"
)

for (package.name in package.name){
  tryCatch(
  {
  if (!require(package.name, character.only = TRUE)){ 
    install.packages(package.name, character.only = TRUE) 
    library(package.name, character.only = TRUE)} else {library(package.name, character.only = TRUE)}
  }, 
    error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}


```

## Set main working directory

```{r working directory}
wd="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/"
setwd(wd)

dir.create(paste(wd,"output/", Sys.Date(), sep=""))
setwd(paste(wd, "output/",Sys.Date(), sep="") )

```

# Load differential expression data from DESeq2

```{r}

l_de=readRDS(paste0(wd, "data/crcpm_a5data/wts/l_a5_de_wts.rds"))

```

# load pathways

```{r}

l_gspathway=readRDS(paste0(wd, "data/genesets/collated/", "l_gspathway.RData"))
l_gspathway=l_gspathway[grepl("HALLMARK", names(l_gspathway))]

# prepare for GSEA
df_genesets=l_gspathway %>% unlist %>% data.frame
df_genesets$gs_name=str_replace_all(rownames(df_genesets), "(?<=\\w)\\d*$", "")
df_genesets$gs_name %>% table
colnames(df_genesets)[1]="gene_symbol"

df_genesets=df_genesets[,c(2,1)]
df_genesets$gs_name=str_replace_all(df_genesets$gs_name, "HALLMARK", "")

```

# GSEA
Run the GSEA analysis with hallmark from msigbr

https://rdrr.io/github/aiminy/ModifiedclusterProfiler/f/vignettes/clusterProfiler.Rmd
https://cran.r-project.org/web/packages/msigdbr/vignettes/msigdbr-intro.html

## Compute GSEA 
Store into a list so that dont have to keep redoing

```{r run GSEA}

l_GSEA=list() # create list to store GSEA data

for(i.comb in names(l_de)){
  
    df_de=l_de[[i.comb]]
    
    # load de data
    original_gene_list <- df_de$log2FoldChange
    
    # cleaning
    names(original_gene_list) <- rownames(df_de) # name the vector
    gene_list<-na.omit(original_gene_list) # omit any NA values 
    gene_list=gene_list[is.finite(gene_list)]
    gene_list = sort(gene_list, decreasing = TRUE) # sort the list in decreasing order (required for clusterProfiler)
    
    # GSEA function - takes into consideration the magnitude of change too!
    gse=GSEA(gene_list, TERM2GENE=df_genesets, verbose=T, pvalueCutoff = 1, pAdjustMethod = "fdr")
    
    l_GSEA[[i.comb]]=gse
    
    gseaplot(gse, geneSetID = 1, by = "runningScore", title = gse$Description[1])
    
}

# save the list for subsequent work
# saveRDS(l_GSEA, file=paste0(wd, "data/", paste0("l_GSEA.RData")))

```

### visualization
#### with dot plots

```{r}

pdf(paste(Sys.Date(), "GSEA_dotplot_a5wts.pdf"), height=4, width=5)

for(i.comb in names(l_GSEA)){
  
  # prepare data
  df_GSEA=l_GSEA[[i.comb]] %>% data.frame
  
  if(nrow(df_GSEA)==0){next()} # remove comparisons without any pathway identified
  
  # cleaning
  df_GSEA$ID=str_replace(df_GSEA$ID, "HALLMARK ", "")
  df_GSEA$ID=str_replace_all(df_GSEA$ID, "_", " ")
  df_GSEA$ID=factor(df_GSEA$ID, levels=unique(df_GSEA$ID[order(df_GSEA$enrichmentScore)]))
  df_GSEA=subset(df_GSEA, !grepl("prolif",df_GSEA$ID))
  
  
  plot_gseadotplot=ggplot()+
                    geom_point(aes(x=NES, y=ID, size=setSize, alpha=p.adjust<0.05, color=NES<0), data=subset(df_GSEA, abs(df_GSEA$NES)>1), stroke=NA)+
                    # scale_color_gradient(low = "darkgoldenrod1", high = "steelblue4")+
                    scale_size(range=c(3,5))+
                    # theme_alluvial()+
                    scale_color_manual(values=c("orange3", "grey10"))+
                    scale_alpha_manual(values=c(0.7, 1))+
                    guides(alpha="none", size="none", color="none")+
                    labs(y="", x=paste0(str_extract(i.comb, "(?<= vs ).*"),"   |   ", str_extract(i.comb, ".*(?= vs )"),"\nNES"))+
                    geom_vline(xintercept=0, col="black") +
                    theme(legend.position="top") 
  
  print(plot_gseadotplot)
  
}

dev.off()

```

#### with GSEA plots

```{r}
pdf(paste(Sys.Date(),"GSEA.pdf"), height=6, width=5)
for (i in 1:length(gse$Description)){
  gseaplot2(gse, 
            title = (str_replace(str_replace_all(gse$Description[i], "_", " "), "HALLMARK ", "")), 
            geneSetID = i) %>% print
}
dev.off()

```
