---
title: "CRCPM: stromal cluster overlaps"
author: "Joseph J Zhao"
date: "4/12/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
package.name=c(
  # data manipulation
  "tidyverse", "readr", "readxl", "dplyr", "tidyr", "lubridate", "tibble", "plyr", "devtools", "stringr", "stringi", "gtools",
  # parallel processing
  "doParallel", "parallel", "foreach", 
  # bioinformatics
  "DESeq2", "BiocParallel", "clusterProfiler", "singscore", "GSVA", "CMScaller", "EnvStats", "org.Hs.eg.db",
  # survival analysis
  "survminer", "survival", "rstpm2", "survRM2",
  # regression stuff
  "MASS", "splines", "Hmisc", "rms",
  # misc
  "cluster", "ResourceSelection", "digitize",
  # plotting
  "blandr", "pheatmap", "ggrepel", "ggplot2", "corrplot", "ggpubr", "gplots", "grid", "ggVennDiagram"
)

for (package.name in package.name){
  tryCatch(
  {
  if (!require(package.name, character.only = TRUE)){ 
    install.packages(package.name, character.only = TRUE) 
    library(package.name, character.only = TRUE)} else {library(package.name, character.only = TRUE)}
  }, 
    error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

```

## Set main working directory

```{r working directory}
wd="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/"
setwd(wd)

dir.create(paste(wd,"output/", Sys.Date(), sep=""))
setwd(paste(wd, "output/",Sys.Date(), sep="") )

```

# tumor to stroma interaction
- evaluated at patient level
- evaluated at slide level

Patient level - already done in the clustering process

```{r}

path="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/submissions/substudies/crcpm/submissions/ccr/r1/workings"

dsp_crc=readRDS(paste0(path, "/data/dsp/dsp_crc.rds"))

crcpm_dspstromalclusters <- read_csv(paste0(path, "/data/dsp/crcpm_stromalclusters/crcpm_dspstromalclusters_results.csv"))[-1]
df_ref_stromalclusters_perpatient <- read_excel(paste0(path,"/data/reference/df_ref_pm_perpatient_crc.xlsx"))
df_ref_pm_persample <- read_excel(paste0(path,"/data/reference/df_ref_pm_persample_annotated_crc.xlsx"))
df_ref_pm_persample=left_join(df_ref_pm_persample, subset(df_ref_stromalclusters_perpatient, select=c("patientid", grep("SC|peritoneal", colnames(df_ref_stromalclusters_perpatient), value=T))), by="patientid")
df_ref_pm_persample=left_join(df_ref_pm_persample, crcpm_dspstromalclusters, by="dspid")

```

venn diagram of overlaps of SC at pateint level - are there patietns with multiple different SC subtypes? (because some patients have more than 1 sample)
https://github.com/gaospecial/ggVennDiagram

```{r}

df_ref_stromalclusters_perpatient

df_dsp_tumorroi=table(subset(df_ref_pm_persample, df_ref_pm_persample$gccrc=="crc" & df_ref_pm_persample$dsp==1)$patientid, 
                      subset(df_ref_pm_persample, df_ref_pm_persample$gccrc=="crc" & df_ref_pm_persample$dsp==1)$stromal_cluster) %>% as.data.frame.matrix
df_dsp_tumorroi$patientid=rownames(df_dsp_tumorroi)

df_ref_stromalclusters_perpatient=left_join(df_ref_stromalclusters_perpatient, df_dsp_tumorroi, by="patientid")

sctumorroi=c("SC1", "SC2", "SC3")

l_venn=list()

for (i.sctumorroi in sctumorroi){
  
  l_venn[[i.sctumorroi]]=subset(df_ref_stromalclusters_perpatient, !df_ref_stromalclusters_perpatient[, i.sctumorroi]==0)$patientid
  
}


pdf(paste(Sys.Date(), "crcpm_scvenndiagram.pdf"), height=3, width=4)
ggVennDiagram(l_venn, force_upset = F) + scale_fill_gradient(low="grey90",high = "purple4")
ggVennDiagram(l_venn, force_upset = TRUE)
dev.off();dev.off();dev.off()



```

stacked barchart of the proportion of PM PT between SCs

```{r}

df_stackbar=subset(df_ref_pm_persample, df_ref_pm_persample$gccrc=="crc" & df_ref_pm_persample$dsp==1 & df_ref_pm_persample$dsproi=="stroma")
df_stackbar=table(df_stackbar$stromal_cluster, df_stackbar$primarypm) %>% as.data.frame.matrix
df_stackbar=apply(df_stackbar, 1, function(x){x/sum(x)})

df_stackbar=reshape2::melt(df_stackbar)
df_stackbar$Var1=factor(df_stackbar$Var1, levels=c("primary", "peritoneal"))

plot_stackprimarypm=ggplot(aes(fill = Var1, x = Var2, y = value), data=df_stackbar) + 
   geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = c("primary"="dodgerblue3", "peritoneal"="firebrick3")) +
  labs(x="", y="% of ROI", fill="")

pdf(paste(Sys.Date(), "crcpm_scprimarypmstack.pdf"), height=3, width=3.5)
print(plot_stackprimarypm)
dev.off()

```

