---
title: "stromal phenotypes orthogonal validation with lunaphore comet"
author: "Joseph J Zhao"
date: "11/6/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
package.name=c(
  # data manipulation
  "tidyverse", "readr", "readxl", "dplyr", "tidyr", "lubridate", "tibble", "plyr", "stringr", "stringi", "gtools", "maditr", "reshape2",
  # parallel processing
  "doParallel", "parallel", "foreach",
  # bioinformatics
  "DESeq2","preprocessCore", "umap", "DGEobj.utils", "edgeR", "sva", "sf", "RANN", "spatstat",
  # survival analysis
  "survminer", "survival", "rstpm2", "survRM2",
  # regression stuff
  "MASS", "splines",
  # misc
  "cluster", "ResourceSelection", "digitize", "statip", "dad",
  # plotting
  "blandr", "pheatmap", "ggrepel", "ggplot2", "corrplot", "ggpubr", "gplots", "grid", "ggbiplot", "ggsankey", "ggVennDiagram", "EnvStats"
)

for (package.name in package.name){
  tryCatch(
  {
  if (!require(package.name, character.only = TRUE)){ 
    install.packages(package.name, character.only = TRUE) 
    library(package.name, character.only = TRUE, quietly = T)} else {library(package.name, character.only = TRUE, quietly = T)}
  }, 
    error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}


```

## Set main working directory

```{r working directory}
wd="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/"
setwd(wd)

dir.create(paste(wd,"output/", Sys.Date(), sep=""))
setwd(paste(wd, "output/",Sys.Date(), sep="") )

```

import

```{r}

path="C:/Users/jzhao/OneDrive/Research_Cloud/NUH_NCIS/gccrc_pm/submissions/substudies/crcpm/submissions/ccr/r1/workings"
l_lunaphore_crcpm=readRDS(paste0(path,"/data/lunaphore/l_lunaphore_crcpm.rds"))

```

heatmap overview of cell type density

```{r}

lcols=list("primarypm"=c("primary"="dodgerblue3", "peritoneal"="firebrick3"),
            "ROI"=c("stroma"="darkgreen", "tumor"="steelblue3", "TSI"="grey"),
           "distance from TSI"=RColorBrewer::brewer.pal(11, "BrBG"))

colMeta=subset(l_lunaphore_crcpm$df_ref_perroi, select=c("primarypm", "roi_type")) %>% as.data.frame
colMeta$roi_type=str_to_lower(colMeta$roi_type); colMeta$roi_type[colMeta$roi_type=="tumor_boundary"]="TSI"
colnames(colMeta)=c("primarypm", "ROI")
rownames(colMeta)=l_lunaphore_crcpm$df_ref_perroi$roiid
heatmap_lunaphore_perroi=pheatmap(t(l_lunaphore_crcpm$df_celltypedensity_perroi), scale="column")
df_lunaphore_clust=data.frame(lunaphore_clust=cutree(heatmap_lunaphore_perroi$tree_col, k=3))

colMeta=merge(colMeta, df_lunaphore_clust, by="row.names")
rownames(colMeta)=colMeta$Row.names; colMeta$Row.names=NULL

heatmap_lunaphore_perroi=pheatmap(t(l_lunaphore_crcpm$df_celltypedensity_perroi), 
         scale="column",
         
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",
         annotation_colors = lcols,
         treeheight_row = 0,
         # treeheight_col=200,
         show_colnames = F,
         annotation_col = colMeta)

pdf(paste(Sys.Date(),paste0("_lunaphore_heatmapperroi.pdf")), height=3.5, width=10)
heatmap_lunaphore_perroi
dev.off()

# include lunaphore_clust into original perroi reference sheet
df_lunaphore_clust

```

DefineSC1 vs SC2 like clusters


```{r}

df_lunaphore_clust$roiid=rownames(df_lunaphore_clust)
l_lunaphore_crcpm$df_ref_perroi=left_join(l_lunaphore_crcpm$df_ref_perroi, df_lunaphore_clust, by="roiid")

# write.csv(l_lunaphore_crcpm$df_ref_perroi, "df_per_roi_lunaphoreclust.csv")

# ignore this

df_lunaphoreclust_perslide=table(subset(l_lunaphore_crcpm$df_ref_perroi, l_lunaphore_crcpm$df_ref_perroi$roi_type=="Stroma")$primarypm, subset(l_lunaphore_crcpm$df_ref_perroi, l_lunaphore_crcpm$df_ref_perroi$roi_type=="Stroma")$lunaphore_clust) %>% as.data.frame.matrix()
df_lunaphoreclust_perslide$"3"=NULL
chisq=chisq.test(df_lunaphoreclust_perslide, correct=F)
colnames(df_lunaphoreclust_perslide)=c("SC1-like", "SC2-like")

# check the breakdown of SC1 vs SC2 like stroma across PTPM - note that there are more primary samples to begin with

df_lunaphoreclust_perslide_melt=melt(t(apply(df_lunaphoreclust_perslide, 2, function(x){x/sum(x)}) ))

colnames(df_lunaphoreclust_perslide_melt)=c("lunaphore_clust", "primarypm", "prop")
df_lunaphoreclust_perslide_melt$primarypm=factor(df_lunaphoreclust_perslide_melt$primarypm, levels=c("primary", "peritoneal"))
df_lunaphoreclust_perslide_melt$lunaphore_clust=factor(df_lunaphoreclust_perslide_melt$lunaphore_clust, levels=c("SC1-like", "SC2-like"))

plot_lunaphoreclustprimarypm_stackedbarchart=ggplot(aes(x=lunaphore_clust, y=prop, fill=primarypm), data=df_lunaphoreclust_perslide_melt) +
  theme_classic2()+
  labs(fill="", x="", y="% of ROI")+
  scale_fill_manual(values = c("dodgerblue3", "firebrick3"))+
    geom_bar(position="fill", stat="identity")

pdf(paste(Sys.Date(),paste0("_lunaphore_clust_stackedbarchart.pdf")), height=3, width=3.5)
print(plot_lunaphoreclustprimarypm_stackedbarchart)
dev.off()


```

TSI distance vs stromal phenotypes
- check if there is any correlation with TSI - as maybe SC2-like stroma is simply closer to tumor and hence is more pro-tumor

```{r}

df_temp=l_lunaphore_crcpm$df_ref_perroi

df_temp$lunaphore_clust[df_temp$lunaphore_clust==1]="SC1-like"
df_temp$lunaphore_clust[df_temp$lunaphore_clust==2]="SC2-like"

plot_stromalcluster_distTSI=ggplot(aes(x=lunaphore_clust, y=dist_tsi/0.23), data=subset(df_temp, df_temp$roi_type=="Stroma" & !df_temp$lunaphore_clust=="3")) + 
  geom_boxplot(width=0.5) +
  labs(x="", y="distance from TSI, μm") +
  stat_compare_means(method="t.test") +
  facet_wrap(~primarypm)+
  stat_n_text()

pdf(paste(Sys.Date(),paste0("_lunaphore_plot_stromalcluster_distTSI.pdf")), height=3, width=4.5)
print(plot_stromalcluster_distTSI)
dev.off()

```

Evaluate spatial distribution
https://www.paulamoraga.com/book-spatial/complete-spatial-randomness.html
Analysing-spatial-point-patterns-in-R.pdf

```{r}

for (i.celltype in c("T cells", "B cells", "Fibroblasts", "CD163 macrophages", "CD68 macrophages", "Tumor cells")){
  for (i.roiid in unique(l_lunaphore_crcpm$df_ref_percell$roiid)){
  
  
    df_temp_cell=subset(l_lunaphore_crcpm$df_ref_percell, l_lunaphore_crcpm$df_ref_percell$roiid==i.roiid & !is.na(l_lunaphore_crcpm$df_ref_percell$celltype) & l_lunaphore_crcpm$df_ref_percell$celltype==i.celltype)

    if (nrow(df_temp_cell)<10){next()}
    
    roi=ppp(df_temp_cell$X_position, df_temp_cell$Y_position, c(0, 2048), c(0, 2048))
    
    # spatial randomness evaluation
    Q=quadratcount(roi, nx = 4, ny = 3)
    # quadrat.test(Q, alternative = "two.sided") # is the distribution random
    p_cluster=quadrat.test(Q, alternative = "clustered") # is the distribution clustered
    
    l_lunaphore_crcpm$df_ref_perroi[[paste0("p_clust_", i.celltype)]][l_lunaphore_crcpm$df_ref_perroi$roiid==i.roiid]=p_cluster$p.value[[1]]
    
    # neaarest nieghtbour distance
    nndist=mean(nndist(roi))
    
    l_lunaphore_crcpm$df_ref_perroi[[paste0("nndist_", i.celltype)]][l_lunaphore_crcpm$df_ref_perroi$roiid==i.roiid]=nndist*0.23

    if (p_cluster$p.value[[1]]>0.05 & nrow(df_temp_cell)>100){
      plot(density(roi), main=""); plot(roi, add=T, pch=19, size=0.75, col=lcols[[i.celltype]])
    }
  }
  
  l_lunaphore_crcpm$df_ref_perroi[[paste0("p_clust_", i.celltype)]]=p.adjust(l_lunaphore_crcpm$df_ref_perroi[[paste0("p_clust_", i.celltype)]], method = "bonferroni")
  
}

df_temp=subset(l_lunaphore_crcpm$df_ref_perroi, l_lunaphore_crcpm$df_ref_perroi$roi_type=="Stroma" & !l_lunaphore_crcpm$df_ref_perroi$lunaphore_clust==3)
df_temp$lunaphore_clust=ifelse(df_temp$lunaphore_clust==1, "SC1-like", "SC2-like")

df_temp_melt=reshape2::melt(subset(df_temp, select=c("primarypm", "roi_type", "lunaphore_clust", "nndist_T cells", "nndist_B cells", "nndist_Fibroblasts", "nndist_CD163 macrophages", "nndist_CD68 macrophages", "nndist_Tumor cells")))
df_temp_melt$primarypm=factor(df_temp_melt$primarypm, levels=c("primary", "peritoneal"))
df_temp_melt$variable=str_replace_all(df_temp_melt$variable, "nndist_", "")

# between sc12 vs ptpm

plot_nndistptpm=ggplot(aes(y=value, x=lunaphore_clust, color=lunaphore_clust), data=subset(df_temp_melt, !df_temp_melt$variable=="Tumor cells")) +
  ggbeeswarm::geom_beeswarm(cex=2, size=0.75) +
  stat_n_text() +
  facet_wrap(~primarypm+variable, nrow = 2) +
  scale_y_log10()+
  guides(color="none")+
  scale_color_manual(values=c("goldenrod3", "darkgreen"))+
  stat_compare_means(method="t.test", size=4)+
  labs(x="", y="average nearest neighbour\ndistance per ROI, μm")

pdf(paste(Sys.Date(),paste0("_lunaphore_nndistscclust.pdf")), height=6, width=9)
print(plot_nndistptpm)
dev.off()

# ptpm only
plot_nndistptpm=ggplot(aes(y=value, x=primarypm, color=primarypm), data=subset(df_temp_melt, !df_temp_melt$variable=="Tumor cells")) +
  ggbeeswarm::geom_beeswarm(cex=2, size=0.75) +
  stat_n_text() +
  facet_wrap(~variable, ncol=1) +
  scale_y_log10()+
  guides(color="none")+
  scale_color_manual(values=c("dodgerblue3", "firebrick3"))+
  stat_compare_means(method="t.test", size=4)+
  labs(x="", y="average nearest neighbour\ndistance per ROI, μm")

pdf(paste(Sys.Date(),paste0("_lunaphore_nndistptpm.pdf")), height=8, width=2.5)
print(plot_nndistptpm)
dev.off()

```

print out some of the images for review - manual selection

```{r}

lcols=c(RColorBrewer::brewer.pal(6, "Dark2"))
names(lcols)=c("B cells", "CD163 macrophages", "CD68 macrophages","Fibroblasts",  "T cells", "Tumor cells")

selected_roi=c("CRC_S19-008963_F-5_A09", "CRC_S17-4025_3-2_A11")

pdf(paste(Sys.Date(),paste0("_lunaphore_roiplots.pdf")), height=3, width=4.5)

for (i.roiid in selected_roi){
  
  df_temp=subset(l_lunaphore_crcpm$df_ref_percell, l_lunaphore_crcpm$df_ref_percell$roiid==i.roiid & !is.na(l_lunaphore_crcpm$df_ref_percell$celltype))
  
  plot_individualroi=ggplot() + 
    theme_void()+
    geom_point(aes(x=X_position, y=Y_position, color=celltype), data=df_temp) +
    labs(color="")+
    scale_color_manual(values=lcols)
  
  print(plot_individualroi)
  
}

dev.off()

selected_roi=c("CRC_S19-008963_F-5_A09", "CRC_S17-4025_3-2_A11")

pdf(paste(Sys.Date(),paste0("_lunaphore_roicelltypedensityplot.pdf")), height=4.5, width=4.5)

for (i.roiid in selected_roi){
  for (i.celltype in c("T cells", "B cells", "Fibroblasts", "CD163 macrophages", "CD68 macrophages", "Tumor cells")[c(1,3)]){
  
  
    df_temp_cell=subset(l_lunaphore_crcpm$df_ref_percell, l_lunaphore_crcpm$df_ref_percell$celltype==i.celltype & l_lunaphore_crcpm$df_ref_percell$roiid==i.roiid)
  
    roi=ppp(df_temp_cell$X_position, df_temp_cell$Y_position, c(0, 2048), c(0, 2048))
    
  plot(density(roi), main=paste0(i.roiid, "\n",i.celltype)); plot(roi, add=T, pch=19, size=0.5, col="black")
    
  }
}

dev.off()

```

